---
title: "Time series in GGplot2"
output: html_notebook
author: suresh
---
In other note, we have plotted expression profile as bar plots for replicates. In this note, we will plot time series gene expression data.  We do not have replicates as the other note does. We have two conditions here (treated and untreated) for time 5 points. We have 50 observations per time point per condition. Hence we have 500 observations (50 values x 2 conditions x 5 times points=500). One thing what users would like to see is expression levels before and afer or normal vs tumor. Let us connect two values by a line. So that we wknow which is what. We can represent the data horizontally or vertically.

### Let us do this
#### Let us simulate a dataframe with 500 values as I explained above.
```{r}
x <- data.frame(replicate(10, sample(0:500, 50, replace=TRUE)))
rownames(x) <- paste("gene", c(1:nrow(x)), sep="")
colnames(x)[seq(1,ncol(x), 2)] <- paste("treated.Month", c(1:(ncol(x)/2)), sep="")
colnames(x)[seq(2,ncol(x), 2)] <- paste("untreated.Month", c(1:(ncol(x)/2)), sep="")
x$id=row.names(x)
```

#### Now let us convert the data from wide to long format
```{r}
library(tidyverse)
gx=gather(x,"treatment","Expression",-id)
```

#### Let us create two more columns after splitting treatment column.
```{r}
library(stringr)
gx[,c("condition","time")]=str_split_fixed(gx$treatment,"\\.",2)
```
#### Convert the new columns to factor
```{r}
gx$time=as.factor(gx$time)
gx$condition=as.factor(gx$condition)
```
#### Take any 6 genes. In this example, we take result of head.
```{r}
gx6=subset(gx,gx$id %in% head(gx$id) )
```
#### Now plot genes vertically
```{r}
library(ggplot2)
ggplot(gx6, aes( id, Expression,  color = condition)) +
    geom_point(size = 4) +
    geom_line(aes(group = id), color = "brown", size = 1) +
    facet_wrap(~ time, nrow = 1, ncol = 5) +
    theme_bw() +
    theme(
          axis.text.x = element_text(angle = 90, size = 17),
          axis.text.y = element_text(size = 17),
          axis.title.x = element_blank(),
          strip.text.x = element_text(size = 15)
          )
```

#### Now plot genes horizontally
```{r}
ggplot(gx6, aes(Expression, id, color = condition)) +
    geom_point(size = 4) +
    geom_line(aes(group = id), color = "brown", size = 1) +
    facet_wrap(~ time, nrow = 1, ncol = 5) +
    theme_bw() +
    theme(
          axis.text.x = element_text(angle = 90, size = 17),
          axis.text.y = element_text(size = 17),
          axis.title.x = element_blank(),
          strip.text.x = element_text(size = 15)
          )

```